Heatwaves Exacerbate Inequalities in Mental Health

### 2020-12-17 doc/temp-sentiment.tex
Latest manuscript edits, comments, text, etc from co-working session Thursday 17Dec2020. Reflects new ABT (And But Therefore) framing and paper re-organization.

\title{Assessing Vulnerability to Mental Health Impacts of Weather}

@@

\begin{abstract}
Weather can affect people’s mood and well-being. Traditional studies exploring the effects of weather on mental health require significant amounts of time to plan, fund and execute.  Continuous in situ data streams from social media, newspapers, and other big data sources offer an opportunity to examine how people perceive and respond to environmental conditions. Recent work demonstrates that prevailing weather conditions can affect people’s sentiment as expressed on Twitter and other social media.  Studies to date however, have modeled the weather as having the same effect on sentiment across all locations and individuals (i.e., in aggregate).  In this study, we explore how weather affects individual expressed sentiment (via Twitter) across different weather gradients and locations. We analyzed the sentiment from a quarter of a billion geolocated Tweets from 2009 to 2019, using multiple measures of sentiment, overlaid with data about prevailing weather conditions, as well as local land cover, income levels, and demographic characteristics in the vicinity of Twitter users.  We find that weather affects people differently across a range of incomes, with people in higher income brackets generally happier with increasing temperature, while people in lower income brackets are less happy with increasing temperature.  This project extends existing research about the relationship between temperature and mood (sentiment) by examining how varying climatic conditions across different geographies and socio-economic groups are correlated with expressed sentiment. This project brings new insights into how humans perceive and respond to environmental shocks, with important implications for economics, policy and adaptation and resilience planning.

@@

\section{Introduction}
This work builds on existing work by \citep{baylis_weather_2018}. 
Consider going through our paper and using the ABT method for ensuring we are building up a logic string, arguments, why we're doing this, what it answers, etc.

@@

Changes in climate result in fluctuations in the frequency, intensity, spatial extent, duration, and timing of weather resulting in unprecedented climate extremes . Changing climate is putting pressures on environmental and social system health and necessitating changes in how we respond to environmental conditions and anticipate (and plan for) future conditions. (cite IPCC report on health and environment, also cite the Lancet report on climate and health).

@@

Although climate change is an international phenomenon experienced in every country, its discussions vary from country to country. According to Vu and others \cite{vu_nationalizing_2019}, a country’s climate severity, economic status and governance determine variations in its media discussions. For example, Park and others \cite{park_mood_2013} \href{https://pdfs.semanticscholar.org/b282/feb759e57530b115dcc4bb080f96598a5246.pdf}{(link} demonstrated that many people living in a state with higher mean temperature do express more positive emotions on Twitter than those living in colder states in the US. Schmidt and others \cite{schmidt_media_2013} posit that climate severity and environmental factors related to carbon dependency influences the amount of coverage global warming receives in the media in countries like USA, Australia and Germany. Studies comparing how different countries portray climate change have been widely conducted in USA, UK, France and the Netherlands \cite{vu_nationalizing_2019}. Such comparisons are mainly between developed countries with developing countries outside the focus.  Schafer and O’Neil \cite{schafer_what_2013} advance the need for academic scholars to investigate transnational contexts within climate communication. Wealthier countries with comparatively higher GDP are likely to frame climate change as a political issue as financial resources exist for exploring research on climate change. Conversely, coverage of climate change news from poorer developing countries focus mainly on international relations. Thus, the effects of national macro economic variables that affects a country’s sociopolitical and economic development such as GDP reflects a country’s governance system in media coverage \cite{vu_what_2018}. Cultural theorists assert that, understanding such differential variations depicts how different entities interpret danger and respond to risk \cite{tansey_cultural_1999}. Inclusion of developing countries in scientific research provides an avenue for international support to such countries in responding to the effects of climate change.Yet studies attempting to explain cross-national variation in climate change public opinion is limited \cite{knight_public_2016}.

@@

Building on previous findings from Baylis and others, we hypothesize that local ambient weather conditions have a strong connection how people are feeling as expressed in the sentiment of text messages posted on Twitter [H1]. Second, we hypothesize that the weather-sentiment link will be stronger in low-income areas [H2]. Third, we hypothesize that people living in areas with less impervious surfaces and more vegetative tree cover will exhibit higher sentiment scores [H3].

\section{Data}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{../res/Boston_Map.png}
  \caption{Here is a map of all tweets in boston, overlaid with income by census block}
  \label{fig:timeseries}
\end{figure}

\section{Methods}
\subsection{Text Data Collection}
Text data used for this project come from Twitter via the University of Vermont’s (UVM) agreement with Twitter to access its streaming API - colloquially referred to as the Decahose - which provides public access to a random 10\% of all public messages on Twitter. The UVM special agreement with Twitter allows for access to this data for research and analysis purposes and we have complied with all the terms of service for Twitter and UVM. 
The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform. We used all geolocated Tweets from the Decahose for the period 2012-2016 with a bounding box polygon filter for India, Western Europe and the United States of America. We only considered users’ original content, thus did not include retweets in our analysis.

@@

\subsection{Income \& Race}
We used data from the US Census at the Census Block level to estimate the

@@

>>>>

\title{Heatwaves Exacerbate Inequalities in Mental Health}

@@

\begin{abstract}
Weather can affect people’s mood and well-being. Traditional studies exploring the effects of weather on mental health require significant resource investment (e.g., time, funding, execution).  Continuous in situ data streams from social media, newspapers, and other big data sources offer an opportunity to examine how people perceive and respond to environmental conditions. Recent work demonstrates that prevailing weather conditions can affect people’s sentiment as expressed on Twitter and other social media.  Studies to date however, have modeled the weather as having the same effect on sentiment across all locations and individuals (i.e., in aggregate).  In this study, we explore how weather affects individual expressed sentiment (via Twitter) across different weather gradients and locations. We analyzed the sentiment from a quarter of a billion geolocated Tweets from 2009 to 2019, using multiple measures of sentiment, overlaid with data about prevailing weather conditions, as well as local land cover, income levels, and demographic characteristics in the vicinity of Twitter users.  We find that weather affects people differently across a range of incomes, with people in higher income brackets generally happier with increasing temperature, while people in lower income brackets are less happy with increasing temperature.  This project extends existing research about the relationship between temperature and mood (sentiment) by examining how varying climatic conditions across different geographies and socio-economic groups are correlated with expressed sentiment. This project brings new insights into how humans perceive and respond to environmental shocks, with important implications for economics, policy and adaptation and resilience planning.

@@

%Cite this new paper: https://doi.org/10.1016/S2542-5196(20)30251-5

\section{Introduction}

% Organize as follows:

% Heatwaves are known to worsen mental health
% AND mental health is strongly associated with income
% AND wealthier better able to cope with heatwaves 
% BUT we know little about the interaction between heatwaves, mental health and income
% THEREFORE: we did this really novel study.

This work builds on existing work by \citep{baylis_weather_2018}. 
Consider going through our paper and using the ABT method for ensuring we are building up a logic string, arguments, why we're doing this, what it answers, etc.

Research has demonstrated strong linkages between ambient environmental conditions and mental health, with extreme temperatures more frequently associated with poorer overall mental health. Mental health is strongly associated with income with people in lower income levels frequently experiencing higher stress levels and less capable of finding help for coping with these stresses. Additionally, results from research on how people cope with environmental temperature extremes indicate that wealthier are better able to cope with heatwaves. Furthermore, previous studies examining relationships between environmental extremes and health sometimes aggregate data at scales too coarse to examine whether income may be an ameliorating factor. Yet we know relatively little about the interaction between temperature, mental health and income. This research examines the interactions between these three using relatively fine-scale, publicly available data. This study expands on existing temperature and sentiment - one measure of mental health - research by examining finer-scale data and incorporating income data to examine correlations between extreme temperatures and sentiment.

\section{Background}

@@

@@

Although climate change is an international phenomenon experienced in every country, its discussions vary from country to country. According to Vu and others \cite{vu_nationalizing_2019}, a country’s climate severity, economic status and governance determine variations in its media discussions. For example, Park and others \cite{park_mood_2013} \href{https://pdfs.semanticscholar.org/b282/feb759e57530b115dcc4bb080f96598a5246.pdf}{(link)} demonstrated that many people living in a state with higher mean temperature do express more positive emotions on Twitter than those living in colder states in the US. Schmidt and others \cite{schmidt_media_2013} posit that climate severity and environmental factors related to carbon dependency influences the amount of coverage global warming receives in the media in countries like USA, Australia and Germany. Studies comparing how different countries portray climate change have been widely conducted in USA, UK, France and the Netherlands \cite{vu_nationalizing_2019}. Such comparisons are mainly between developed countries with developing countries outside the focus.  Schafer and O’Neil \cite{schafer_what_2013} advance the need for academic scholars to investigate transnational contexts within climate communication. Wealthier countries with comparatively higher GDP are likely to frame climate change as a political issue as financial resources exist for exploring research on climate change. Conversely, coverage of climate change news from poorer developing countries focus mainly on international relations. Thus, the effects of national macro economic variables that affects a country’s sociopolitical and economic development such as GDP reflects a country’s governance system in media coverage \cite{vu_what_2018}. Cultural theorists assert that, understanding such differential variations depicts how different entities interpret danger and respond to risk \cite{tansey_cultural_1999}. Inclusion of developing countries in scientific research provides an avenue for international support to such countries in responding to the effects of climate change.Yet studies attempting to explain cross-national variation in climate change public opinion is limited \cite{knight_public_2016}.

@@

Building on previous findings from Baylis and others, we hypothesize that local ambient weather conditions have a strong connection how people are feeling as expressed in the sentiment of text messages posted on Twitter [H1]. Second, we hypothesize that the weather-sentiment link will be stronger in low-income areas. That is to say, people in lower income areas will be have lower sentiment scores than those in wealthier areas. [H2]. Third, we hypothesize that people living in areas with less impervious surfaces and more vegetative tree cover will exhibit higher sentiment scores [H3].

\section{Data}
A proposed list of tables and figures:
Figures for location of USA Tweets (to show aggregation and coverage).
Figures of income blocks and temperature extremes.
Table 1. Variable descriptions for what's included in our models.
Table 2. Descriptive statistics for datasets/variables we include.
Table 3. Model results with variables included in the models coefficients, SEs, p-values, sample size.
Appendix table. Sentiment scores across all types we considered.
\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{../res/Boston_Map.png}
  \caption{Here is a map of all tweets in boston, overlaid with income by census block}
  \label{fig:timeseries}
\end{figure}

\section{Methods}
\subsection{Text Data}
Text data used for this project come from Twitter via the University of Vermont’s (UVM) agreement with Twitter to access its streaming API - colloquially referred to as the Decahose - which provides public access to a random 10\% of all public messages on Twitter. The UVM special agreement with Twitter allows for access to this data for research and analysis purposes and we have complied with all the terms of service for Twitter and UVM. 
The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform. We used all geolocated Tweets from the Decahose for the period 2012-2016 with a bounding box polygon filter for India, Western Europe and the United States of America. We only considered users’ original content, thus did not include retweets in our analysis.

@@

\subsection{Income}
We used data from the US Census at the Census Block level to estimate the

@@

From Intro/Background
Changes in climate result in fluctuations in the frequency, intensity, spatial extent, duration, and timing of weather resulting in unprecedented climate extremes . Changing climate is putting pressures on environmental and social system health and necessitating changes in how we respond to environmental conditions and anticipate (and plan for) future conditions. (cite IPCC report on health and environment, also cite the Lancet report on climate and health).
